---
title: "Exploratory Data Analysis of Dimonds data set"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this report I'm going to work with the data set, comtaining the prices and other attributes of almost 54,000 diamonds. The data set comes with ggplot2 library.

```{r first data exploration}
library(ggplot2)
dim(diamonds)
str(diamonds)
```
  
Notes:  
* There are 53940 observations in the data set, 10 variables, 3 ordered factors.  
* According to http://www.lumeradiamonds.com/, the less body color in a white diamond, the more true color it will reflect, and thus the greater its value - D.    

## Diamonds price histogram

The diamond price histogram from the diamonds data set

```{r pressure}
qplot(data = diamonds, x = price, binwidth = 250,
      color = I('black'), fill = I('#F79420'),
      xlab = "Diamonds' price in $",
      ylab = "Count of diamonds") + 
  scale_x_continuous(breaks = seq(0,20000, 2000))
```
  
Notes:  
* The diamond price distribution is positively skewed, has a peak (mode) around 800$ and long right tail.  
* The median of the distribution is shifted to the left and equal to 1838$. Mean is 3170$.  
* Min diamond price is 357$ and max is 18690$.  

##Diamonds price explorations
```{r}
#How many diamonds cost less than $500?
NROW(subset(diamonds,price<500))
#How many diamonds cost less than $250?
NROW(subset(diamonds,price<250))
#How many diamonds cost $15000 and more?
NROW(subset(diamonds,price>=15000))
```
##Exploring the largest peak in the price histogram
```{r}
library(gridExtra)
q <- qplot(data = diamonds, x = price, binwidth = 1,
      color = I('black'), fill = I('#F79420'),
      xlab = "Diamonds' price in $",
      ylab = "Count") 
q2 <- q + scale_x_continuous(breaks = seq(0,3000, 100), limits = c(350,2000))
ggsave('priceHistogram1.png')
q3 <- q + scale_x_continuous(breaks = seq(0,3000, 50), limits = c(575,825))
ggsave('priceHistogram2.png')
#exploring the price range with no diamonds
q4 <- q + scale_x_continuous(breaks = seq(0,3000, 50), limits = c(1400,1600))
ggsave('priceHistogram3.png')
suppressWarnings(grid.arrange(q2,q3,q4,ncol=1))
ggsave('priceHistogram.png')
```

Notes:  
* The largest peak of diamonds' price distribution is in the range from 500$ to 850$.  
* Most  of the diamonds have the price a bit more than 600$. To be clarified further.  
* There is a price range between 1450$ and 1550$ when we don't have any diamond.  

Top 3 diamonds' prices:  
```{r}
head(sort(table(diamonds$price), decreasing = T),n = 3)
```
1.  132 diamonds with the price 605$  
2.  127 diamonds for 802$  
3.  126 diamonds for 625$  

Exploring the price range without diamonds
```{r}
table(subset(diamonds$price, diamonds$price > 1450 & diamonds$price < 1550))
```
There are no diamonds in price range from 1455$ till 1545$

##The histogram of diamond prices broken up by cut
```{r}
q + facet_wrap(~cut, scales = "free_y")

```
##The histogram of price per carat and by cut facet
```{r}
qplot(data = diamonds, x = price/carat, binwidth = 0.025,
      xlab = "Price per carat, $",
      ylab = "Counts") +
  facet_wrap(~cut, scales = "free_y") +
  scale_x_log10()

```

Notes:  
* there are still the diamonds with the not-ideal cut and huge price  
* there not many "cheap" diamonds with fair cut, but better is cut, more "cheap" diamonds

```{r}
by(diamonds$price,diamonds$cut,summary,digits = max(getOption('digits')))
```

Notes:  
* The lowest price have diamonds with Premium and Ideal cuts  
* The highest price have diamonds with Premium cut
* The lowest median price have diamonds with Ideal cut (surprise)  

##Boxplot by clarity, color and cut
Visualizing the box plots, I would like to understand how the price per carat depends from the clarity, color and cut. I expect to see that better clarity (IF - good vs. I1 - bad), better color (D - good vs. J - bad) and better cut causes higher price median.
```{r}
b1 <- qplot(data = diamonds, x = clarity, y = price/carat,
      xlab = 'Clarity',
      ylab = 'Price',
      geom = 'boxplot',
      fill=clarity) +
  coord_cartesian(ylim=c(350,18700))
ggsave("boxplot-price-clarity.png")
b2 <- qplot(data = diamonds, x = color, y = price/carat,
      xlab = 'Color',
      ylab = 'Price',
      geom = 'boxplot',
      fill=color) +
  coord_cartesian(ylim=c(350,18700))
ggsave("boxplot-price-color.png")
b3 <- qplot(data = diamonds, x = cut, y = price/carat,
      xlab = 'Cut',
      ylab = 'Price',
      geom = 'boxplot',
      fill=cut) +
  coord_cartesian(ylim=c(350,18700))
b1
b2
b3
ggsave("boxplot-price-cut.png")
```

Against expectations, looking on price - clarity boxplot, median price for the one of the worst clarity level SI2 is higher than IF clarity. At the same time, I can see that the most expensive diamonds with price per carat higher than 15000$ has just IF - best - clarity level and the maximum price - clarity has relatively high linear correlation.  

Let's add color on the same plot and we can see something interesting:
```{r}
qplot(data = diamonds, x = clarity, y = price/carat,
      xlab = 'Clarity',
      ylab = 'Price',
      geom = 'boxplot',
      fill=color) +
  coord_cartesian(ylim=c(350,18700))
ggsave("boxplot-price-clarity-color.png")
```

The diamonds with the best color and clarity have the highest price per carat.
Let's check how the plot will change if we add cut instead of color:
```{r}
qplot(data = diamonds, x = clarity, y = price/carat,
      xlab = 'Clarity',
      ylab = 'Price',
      geom = 'boxplot',
      fill=cut)
ggsave("boxplot-price-clarity-cut.png")
```
The picture is not so obvious for best clarity and color but there is a linear regression of price clarity - cut

##Measures of central tendency
```{r}
by(diamonds$price,diamonds$color,summary,digits = max(getOption('digits')))
```

* The price range for the middle 50% of diamonds with color D:  
  * first quartile (25%): 911$  
  * third quartile (75%): 4213.5$  
* The price range for the middle 50% of diamonds with color J:  
  * first quartile (25%): 1860.5$    
  * third quartile (75%): 7695$  
* IQR for diamonds with the best color: 3302.50  
* IQR for diamonds with the worst color: 5834.50    

##Diamond's size exploration
Carat Frequency Polygon
```{r}
qplot(data = diamonds, x = carat, y = ..count.., 
      binwidth = 0.01,
      geom = "freqpoly") +
  scale_x_continuous(limits = c(0, 2.2), breaks = seq(0, 2.2, 0.1))

```

Most of the diamonds have the size 0.3 carat

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
